# Chapter 6: Analysis of longitudinal data

*This work includes performing and interpreting analysis of longitudinal data technique. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Table 8.1: RATS Measurements from 16 Subjects

```{r}
library(tidyr); library(dplyr); library(ggplot2)
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
glimpse(RATS)
head(RATS)
tail(RATS)
# Convert data to long form:
RATSL <- pivot_longer(RATS, cols=-c(ID,Group), names_to = "WD",values_to = "Weight")  %>%  mutate(Time = as.integer(substr(WD,3,4))) %>% arrange(Time)
glimpse(RATSL)
head(RATSL)
tail(RATSL)
```
\pagebreak

## Figure 8.1

```{r}
p1 <- ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
p6
```
\pagebreak

## Figure 8.2

```{r}
# Standardise the scores:
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate( stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)
p1 <- ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(name = "standardized weight")
p6
```
\pagebreak

## Figure 8.3

```{r}
# Number of subjects (per group):
n <- 58
# Make a summary data:
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean=mean(Weight), se=sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)
p1 <- ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group))
p2 <- p1 + geom_line() + scale_linetype_manual(values = c(1,2,3))
p3 <- p2 + geom_point(size=3) + scale_shape_manual(values = c(1,2,3))
p4 <- p3 + geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3)
p5 <- p4 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6 <- p5 + theme(legend.position = c(0.8,0.8))
p7 <- p6 + scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
p7
```
\pagebreak

## Figure 8.4

```{r}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + theme(legend.position = c(0.8,0.8))
p5 <- p4 + scale_x_discrete(name = "Time")
# Black & White version:
#p6 <- p5 + scale_fill_grey(start = 0.5, end = 1)
p5
```
\pagebreak

## Figure 8.5

```{r}
# Make a summary data of the post Group Times (1-8)
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
glimpse(RATSL8S)
p1 <- ggplot(RATSL8S, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Times 1-8")
p5
```
\pagebreak

## Figure 8.6

```{r}
# Remove the outlier:
RATSL8S1 <- RATSL8S %>%
  filter(mean < 600)
glimpse(RATSL8S1)
p1 <- ggplot(RATSL8S1, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Times 1-8")
p5
```
\pagebreak

## Table 8.3

```{r}
# Without the outlier, apply Student's t-test, two-sided:
#t.test(mean ~ Group, data = RATSL8S1, var.equal = TRUE)
```
\pagebreak

## Table 8.4

```{r}
# Add the baseline from the original data as a new variable to the summary data:
baseline <- RATS$WD1
RATSL8S2 <- RATSL8S %>%
  mutate(baseline)
# Fit the ANCOVA model and see the results:
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)
summary(fit)
anova(fit)
```
\pagebreak






# Table 9.1: bprs of BPRs Recorded Over a 3-Week Period

```{r, echo=TRUE}
library(tidyr); library(dplyr); library(ggplot2)
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
glimpse(BPRS)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
```
\pagebreak

## Table 9.2

```{r, echo=TRUE}
BPRSL <-  pivot_longer(BPRS, cols=-c(treatment,subject),names_to = "weeks",values_to = "bprs") %>% arrange(weeks)
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))
glimpse(BPRSL)
head(BPRSL); tail(BPRSL)
```
\pagebreak

## Figure 9.1

```{r, echo=TRUE}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p2 <- p1 + geom_text(aes(label = treatment))
p3 <- p2 + scale_x_continuous(name = "week", breaks = seq(0, 60, 10))
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw()
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```
\pagebreak

## Table 9.3

```{r, echo=TRUE}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
```
\pagebreak

## Figure 9.2

```{r, echo=TRUE}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p2 <- p1 + geom_line()
p3 <- p2 + scale_x_continuous(name = "week", breaks = seq(0, 3,1))
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw() + theme(legend.position = "top")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```
\pagebreak

## Figure 9.3

```{r, echo=TRUE, fig.width=10, fig.height=10}
pairs(BPRS[, 3:11], cex = 0.7)
```
\pagebreak

## Table 9.4

```{r, echo=TRUE}
library("lme4")
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
```
\pagebreak

## Table 9.5

```{r, echo=TRUE}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
anova(BPRS_ref1, BPRS_ref)
```
\pagebreak

## Table 9.6

```{r, echo=TRUE}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
# dummies (in Table) vs summary output: D1 = Group2, D2 = Group3
anova(BPRS_ref1, BPRS_ref2)
```
\pagebreak

## Figure 9.4

```{r, echo=TRUE, fig.width=3, fig.height=4}
Fitted <- fitted(BPRS_ref2)
BPRSL <- BPRSL %>% mutate(Fitted)
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p2 <- p1 + geom_line()
p3 <- p2 + scale_x_continuous(name = "week", breaks = seq(0, 3, 1))
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw() + theme(legend.position = "right") # "none" in the book
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Observed")
graph1 <- p7
p1 <- ggplot(BPRSL, aes(x = week, y = Fitted, group = subject))
p2 <- p1 + geom_line()
p3 <- p2 + scale_x_continuous(name = "Tweel", breaks = seq(0, 3, 1))
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw() + theme(legend.position = "right")
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p7 <- p6 + ggtitle("Fitted")
graph2 <- p7
graph1; graph2
```
\pagebreak

